# US Census Assignment
========================================================
Libraries and functions needed throughout the assignment
```{r message=FALSE}
library("ggplot2")
library("caret")
library("randomForest")
library("tabplot")

# Multiple plot function 
# http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
# 
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

## Import and explore the data

Read in the learning dataset:
```{r}
Sys.setlocale('LC_ALL','C') # because some strings are invalid in the default locale
census_learn <- read.csv("us_census_full/census_income_learn.csv", header=F) # set path accordingly
summary(census_learn)
```
"?" = missing data  
Throughout this assignment, I didn't impute any missing data.  
"Not in universe" = When to include, when not? For example: 
```{r}
# working class
summary(census_learn$V2) # working class
sum(census_learn$V1<10) # younger than 10; they probably never worked  
```
In practice, I'd try to get to know more about these cases (in this case I couldn't open the [link](http://www.bls.census.gov/cps/cpsmain.htm)). Here, for simplicity, I just keep it.

V25 seems to be *instance weight*
```{r}
census_learn <- census_learn[, -25]
colnames(census_learn) <- c('age', 'work_class', 'industry','occupation', 'education', 'wage_ph', 'enroll_edu', 'm_stat', 'industry_major', 'occupation_major', 'race', 'hispanic', 'sex', 'union_member', 'unemployment', 'employment_stat', 'capital_gains', 'capital_losses', 'dividends', 'tax_stat', 'ex_region', 'ex_state', 'household_stat', 'household_summary', 'change_msa', 'change_reg', 'within_reg', 'live_1ya', 'mig_sun', 'for_employer', 'family_und18', 'cob_dad', 'cob_mom', 'cob', 'citizenship', 'own_bus', 'veterans_quest', 'veterans_bnfts', 'weeks', 'year', 'income')
```

**Quick visualisation**
```{r}
# Our missing data will be denoted in the legend as "?"
# I don't like these plots, I don't find them intuitive to interpret but I didn't know how to quickly visualise %missing data. I could've done the same visualisation as for the cleaned dataset, tho.

p <- tablePrepare(census_learn)
for(i in 1:floor(length(census_learn)/5)){
  print(tableplot(p, select=(i*5-4):(i*5)))
}
tableplot(p, select=(length(census_learn)-length(census_learn)%%5 + 1):length(census_learn))
```

```{r}
# grep -v "?" census_income_learn.csv > census_income_learn_nomiss.csv
census_learn_clean <- read.csv("us_census_full/census_income_learn_nomiss.csv", header=F) 
# fraction of the original dataset that remained
dim(census_learn_clean)[1]/dim(census_learn)[1]
summary(census_learn_clean)
# V16 (employment_stat) and V41 (year) the same for all people in the dataset
```


```{r}
census_learn_clean <- census_learn_clean[, -25]
colnames(census_learn_clean) <- c('age', 'work_class', 'industry','occupation', 'education', 'wage_ph', 'enroll_edu', 'm_stat', 'industry_major', 'occupation_major', 'race', 'hispanic', 'sex', 'union_member', 'unemployment', 'employment_stat', 'capital_gains', 'capital_losses', 'dividends', 'tax_stat', 'ex_region', 'ex_state', 'household_stat', 'household_summary', 'change_msa', 'change_reg', 'within_reg', 'live_1ya', 'mig_sun', 'for_employer', 'family_und18', 'cob_dad', 'cob_mom', 'cob', 'citizenship', 'own_bus', 'veterans_quest', 'veterans_bnfts', 'weeks', 'year', 'income')
```

```{r}
sum(sapply(census_learn_clean, is.factor))
# although was obvious from "year"
# factorising...
cont <- c(1, 6, 17, 18, 19, 30, 39)
nominal <- (1:41)[-cont]
census_learn_clean[, nominal] <- lapply(census_learn_clean[, nominal], as.factor)
```

### Exploring continuous attributes
#### Age
```{r}
tmp <- stack(data.frame(census_learn_clean[,1]))
p1 <- ggplot(tmp, aes(x = ind, y = tmp$values)) + stat_boxplot(geom = 'errorbar') + geom_boxplot(outlier.size = 3) + labs(x = "All", y = "Age") + theme_bw()
p2 <- ggplot(tmp, aes(tmp$values)) + geom_histogram(binwidth = 10) + theme_bw() + labs(x = "Age", y = "Count") 
p3 <- ggplot(census_learn_clean, aes(x = income, y = age, fill = income)) + stat_boxplot(geom = 'errorbar') + geom_boxplot(alpha = 0.2, outlier.size = 3) + labs(x = "Income", y = "Age") + theme_bw()
p4 <- ggplot(census_learn_clean, aes(x = age, color = income, fill = income)) + geom_histogram(binwidth = 10, alpha = 0.2, position="identity") + labs(x = "Age", y = "Count") + theme_bw()

multiplot(p1, p3, p2, p4, cols = 2)
```

#### Wage per hour
```{r}
tmp <- stack(data.frame(census_learn_clean[,6]))
p1 <- ggplot(tmp, aes(x = ind, y = tmp$values)) + stat_boxplot(geom = 'errorbar') + geom_boxplot(outlier.size = 3) + labs(x = "All", y = "Wage per hour") + theme_bw()
p2 <- ggplot(tmp, aes(tmp$values)) + geom_histogram(binwidth = 100) + labs(x = "Wage per hour", y= "Count") + theme_bw()
p3 <- ggplot(census_learn_clean, aes(x = income, y = wage_ph, fill = income)) + stat_boxplot(geom = 'errorbar') + geom_boxplot(alpha = 0.2, outlier.size = 3) + labs(x = "Income", y = "Wage per hour") + theme_bw()
p4 <- ggplot(census_learn_clean, aes(x = wage_ph, color = income, fill = income)) +
geom_histogram(binwidth = 100, alpha = 0.2, position="identity") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme_bw() + labs(x = "Wage per hour", y = "Count") 
multiplot(p1, p3, p2, p4, cols=2)
```

```{r}
sum(census_learn_clean[,6]==0)/length(census_learn_clean[,6])
```
Don't include in the model; > 94% earn 0$ per hour, could be missing data

#### Capital gains
```{r}
tmp <- stack(data.frame(census_learn_clean[,17]))
p1 <- ggplot(tmp, aes(x = ind, y = tmp$values)) + stat_boxplot(geom = 'errorbar') + geom_boxplot(outlier.size = 3) + labs(x = "All", y = "Capital gains") + theme_bw()
p2 <- ggplot(tmp, aes(tmp$values)) + geom_histogram(binwidth = 100) + labs(x = "Capital gains", y = "Count") + theme_bw()
p3 <- ggplot(census_learn_clean, aes(x = income, y = capital_gains, fill = income)) + stat_boxplot(geom = 'errorbar') + geom_boxplot(alpha = 0.2, outlier.size = 3) + labs(x = "Income", y = "Capital gains") + theme_bw()
p4 <- ggplot(census_learn_clean, aes(x = capital_gains, color = income, fill = income)) +
geom_histogram(binwidth = 100, alpha = 0.2, position="identity") + theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust = 1))  + labs(x = "Capital gains", y = "Count") 
multiplot(p1, p3, p2, p4, cols=2)
```
```{r}
sum(census_learn_clean[,17]==0)/length(census_learn_clean[,17])
```
Similar as above... Don't include in the model; > 96% have 0 capital gains

#### Capital losses
```{r}
tmp <- stack(data.frame(census_learn_clean[,18]))
p1 <- ggplot(tmp, aes(x = ind, y = tmp$values)) + stat_boxplot(geom = 'errorbar') + geom_boxplot(outlier.size = 3) + labs(x = "All", y = "Capital losses") + theme_bw()
p2 <- ggplot(tmp, aes(tmp$values)) + geom_histogram(binwidth = 100) + labs(x = "Capital losses", y = "Count") + theme_bw()
p3 <- ggplot(census_learn_clean, aes(x = income, y = capital_losses, fill = income)) + stat_boxplot(geom = 'errorbar') + geom_boxplot(alpha = 0.2, outlier.size = 3) + labs(x = "Income", y = "Capital losses") + theme_bw()
p4 <- ggplot(census_learn_clean, aes(x = capital_losses, color = income, fill = income)) +
geom_histogram(binwidth = 100, alpha = 0.2, position="identity") + theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + labs(x = "Capital losses", y = "Count") 
multiplot(p1, p3, p2, p4, cols=2)
```
```{r}
sum(census_learn_clean[,18]==0)/length(census_learn_clean[,18])
```
Similar as above... Don't include in the model; > 98% have 0 capital losses

#### Dividents from stocks
```{r}
tmp <- stack(data.frame(census_learn_clean[,19]))
p1 <- ggplot(tmp, aes(x = ind, y = tmp$values)) + stat_boxplot(geom = 'errorbar') + geom_boxplot(outlier.size = 3) + labs(x = "All", y = "Dividends") + theme_bw()
p2 <- ggplot(tmp, aes(tmp$values)) + geom_histogram(binwidth = 100) + labs(x = "Dividends", y = "Count") + theme_bw()
p3 <- ggplot(census_learn_clean, aes(x = income, y = dividends, fill = income)) + stat_boxplot(geom = 'errorbar') + geom_boxplot(alpha = 0.2, outlier.size = 3) + labs(x = "Income", y = "Dividends") + theme_bw()
p4 <- ggplot(census_learn_clean, aes(x = dividends, color = income, fill = income)) + geom_histogram(binwidth = 100, alpha = 0.2, position="identity") + theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + labs(x = "Dividends", y = "Count")
multiplot(p1, p3, p2, p4, cols=2)
```
```{r}
sum(census_learn_clean[,19]==0)/length(census_learn_clean[,19])
```
Similar as above... Don't include in the model; > 89% have 0 dividends from stocks

#### Number of persons worked for employer
```{r}
tmp <- stack(data.frame(census_learn_clean[,30]))
p1 <- ggplot(tmp, aes(x = ind, y = tmp$values)) + stat_boxplot(geom = 'errorbar') + geom_boxplot(outlier.size = 3) + labs(x = "All", y = "#persons") + theme_bw()
p2 <- ggplot(tmp, aes(tmp$values)) + geom_histogram(binwidth = 1) + labs(x = "#persons", y = "Count") + theme_bw()
p3 <- ggplot(census_learn_clean, aes(x = income, y = for_employer, fill = income)) + stat_boxplot(geom = 'errorbar') + geom_boxplot(alpha = 0.2, outlier.size = 3) + labs(x = "Income", y = "#persons") + theme_bw()
p4 <- ggplot(census_learn_clean, aes(x = for_employer, color = income, fill = income)) + geom_histogram(binwidth = 1, alpha = 0.2, position="identity") + labs(x = "#persons", y = "Count") + theme_bw()
multiplot(p1, p3, p2, p4, cols=2)
```

#### Weeks worked in a year
```{r}
tmp <- stack(data.frame(census_learn_clean[,39]))
p1 <- ggplot(tmp, aes(x = ind, y = tmp$values)) + stat_boxplot(geom = 'errorbar') + geom_boxplot(outlier.size = 3) + labs(x = "All", y = "#weeks") + theme_bw()
p2 <- ggplot(tmp, aes(tmp$values)) + geom_histogram(binwidth = 5) + labs(x = "#weeks", y = "Count") + theme_bw()
p3 <- ggplot(census_learn_clean, aes(x = income, y = weeks, fill = income)) + stat_boxplot(geom = 'errorbar') + geom_boxplot(alpha = 0.2, outlier.size = 3) + labs(x = "Income", y = "#weeks") + theme_bw()
p4 <- ggplot(census_learn_clean, aes(x = weeks, color = income, fill = income)) + geom_histogram(binwidth = 5, alpha = 0.2, position="identity") + labs(x = "#weeks", y = "Count") + theme_bw()
multiplot(p1, p3, p2, p4, cols=2)
```

### Exploring nominal variables

```{r}
for(i in 1:length(nominal)){
print(ggplot(census_learn_clean, aes(factor(census_learn_clean[,nominal[i]]), fill=factor(income))) + theme_bw() + geom_bar(aes(y=..count../sum(..count..))) + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + labs(x = colnames(census_learn_clean)[nominal[i]], y = "Percent"))
}
```

## Modelling the data
```{r}
# age, for_employer, weeks
# work_class, occupation, education, m_stat, race, sex, citizenship
```

### Logistic regression
```{r}
model_glm1 <- glm(income ~ age + for_employer + weeks + work_class + occupation + education + m_stat + race + sex + citizenship, family=binomial(link='logit'), data = census_learn_clean)  
summary(model_glm1)
# occupation46                                                 NA       NA 
# https://www.r-bloggers.com/learn-logistic-regression-and-beyond/
# remove occupation
model_glm2 <- glm(income ~ age + for_employer + weeks + work_class + education + m_stat + race + sex + citizenship, family=binomial(link='logit'), data = census_learn_clean)  
summary(model_glm2)
# remove work_class
model_glm3 <- glm(income ~ age + for_employer + weeks + education + m_stat + race + sex + citizenship, family=binomial(link='logit'), data = census_learn_clean) 
summary(model_glm3)
predict_glm <- predict(model_glm3, census_learn_clean[, -41], type = 'response')
predictions_glm <- ifelse(predict_glm >= 0.5, " 50000+.", " - 50000.") 
sum(predictions_glm==as.character(census_learn_clean$income))/length(census_learn_clean$income)
confusionMatrix(predictions_glm, as.character(census_learn_clean$income))
# warning but a sensible answer
```

### Random forest
```{r}
set.seed(1906)
model_rf <- randomForest(income ~ age + for_employer + weeks + education + m_stat + race + sex + citizenship, data= census_learn_clean)
print(model_rf)
predictions_rf <- predict(model_rf, census_learn_clean[, -41], type = 'class')
confusionMatrix(predictions_rf, census_learn_clean$income)
# warning but a sensible answer
```
## Making predictions
I made predictions on the cleaned test dataset.
```{r}
# grep -v "?" census_income_test.csv > census_income_test_nomiss.csv

census_test_clean <- read.csv("us_census_full/census_income_test_nomiss.csv", header=F) 
census_test_clean <- census_test_clean[, -25]
colnames(census_test_clean) <- c('age', 'work_class', 'industry','occupation', 'education', 'wage_ph', 'enroll_edu', 'm_stat', 'industry_major', 'occupation_major', 'race', 'hispanic', 'sex', 'union_member', 'unemployment', 'employment_stat', 'capital_gains', 'capital_losses', 'dividends', 'tax_stat', 'ex_region', 'ex_state', 'household_stat', 'household_summary', 'change_msa', 'change_reg', 'within_reg', 'live_1ya', 'mig_sun', 'for_employer', 'family_und18', 'cob_dad', 'cob_mom', 'cob', 'citizenship', 'own_bus', 'veterans_quest', 'veterans_bnfts', 'weeks', 'year', 'income')
sum(sapply(census_test_clean, is.factor))
cont <- c(1, 6, 17, 18, 19, 30, 39)
nominal <- (1:41)[-cont]
census_test_clean[, nominal] <- lapply(census_test_clean[, nominal], as.factor)
predictions_test <- predict(model_rf, census_test_clean[, -41], type = 'class')
confusionMatrix(predictions_test, census_test_clean$income)
```
